Subject-specific curves for time series forecasting of smart meter demand

### 2020-04-01 sscts-paper-1-wp.Rmd
Update paper

---
title: "Subject-specific curves for time series forecasting of smart meter demand"
author:
- familyname: Roach
  othernames: Cameron
  address: Monash University
  email: cameron.roach@monash.edu
  correspondingauthor: true
- familyname: Hyndman
  othernames: Rob
  address: Monash University
- familyname: Taieb
  othernames: Souhaib Ben
  address: University of Mons
abstract: Buildings are typically equipped with smart meters to measure demand at regular intervals. Smart meter data for a single building have many uses, such as forecasting and assessing overall building performance. However, when data are available from multiple buildings, smart meter data have additional uses that are rarely explored. For instance, we can explore how different building characteristics influence energy demand. If each building is treated as a random effect and building characteristics are handled as fixed effects, a mixed effects model can be used to estimate how characteristics affect energy usage. In this paper we demonstrate that producing one-day ahead demand predictions for 123 commercial office buildings using mixed models can improve forecasting accuracy. We experiment with random intercept, random intercept and slope, and subject-specific curve mixed models. The predictive performance of the mixed effects models are tested against naive, linear and nonlinear benchmark models fitted to each building separately. Having justified the use of a mixed model framework, we provide an example showing how mixed model frameworks can, when combined with smart meter data and building attributes, be used to carry out scenario analysis. We demonstrate how expected electricity consumption may increase or decrease given a change in building attributes. This research justifies using mixed models to improve forecasting accuracy and to quantify changes in energy consumption under different building configuration scenarios.
keywords: "time series forecasting, mixed-effects models, smart meters, energy, electricity"
wpnumber: 1/19
jelcodes: C10,C14,C52
blind: false
cover: false
toc: false
bibliography: ["library.bib", "packages.bib"]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    # includes:
    #   in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

To the author's knowledge few papers have explored using mixed models in an electricity demand forecasting role. @Brabec2008-jf appears to be closest to this area. In their paper, a nonlinear mixed effects model (NLME) was used to forecast daily gas demand for individual customers. Predictors such as day of week and temperature were treated as random effects. Their NLME model was benchmarked against ARIMAX and ARX approaches. The paper concluded by saying there was no clear winner between the NLME and benchmark models and that both potentially have strengths and weaknesses. Unfortunately, there are few other papers within the energy field that use mixed effects models^[Some papers claim to use mixed models. However, this term is often applied to cases where a combination of models have been used which is different to mixed effects models.] for forecasting.

@@

Few papers have attempted to assess the impact of differences in building characteristics through statistical methods. To the author's knowledge, only a previous paper by @Roach2019-ss has looked into this using mixed effects models. Whereas that paper focused on estimating demand impact profiles for building attributes at different times of the year, this paper focuses on improving forecast accuracy and producing scenario analyses by estimating the expected change in time series conditional on the building attributes.

<!-- It is our suspicion that mixed-models may prove particularly useful when exploring predictor-response relationships close to the extremes of a predictor's domain, where data may be sparse. For example, peak electricity demand usually occurs during temperature extremes (very hot days), but there are relatively few extreme weather days. Using mixed models allows us to use peak demand data from across all buildings to obtain more reliable predictions. This paper not only explores forecast accuracy across all times, but also explores if accuracy is in fact improved during extreme weather events. -->

@@

The main contribution of this paper is to present an approach to forecasting electricity demand for individual buildings using a mixed effects framework. Furthermore, we show how such a model can be used to conduct scenario analysis allowing us to quantify expected energy savings given changes in building attributes. Finally, this paper serves to enrich the literature on forecasting with mixed effects models and using smart meter data.

This paper is structured as follows. Section \@ref(data) describes the data we are working with. Section \@ref(methodology) gives a detailed description of the models we are working with and how they are assessed. Forecasting results and a scenario analysis example are presented in Section \@ref(results). Concluding remarks are given in Section \@ref(conclusion).

@@

The relationship between current temperature and electricity demand is shown in Figure \@ref(fig:temp-wh-scatter) for four buildings. We can see a clear difference in this relationship for the working and non-working periods. Furthermore, this relationship varies between each building. Note that when splines are fitted independently for each building they are very sensitive to outliers or data at domain extremes. This can be seen for building BID0052A at midnight (period 96) where a single observation with relatively high demand at the highest temperature results in the fitted spline increasing rapidly (and possibly unrealistically). This partly motivates our use of mixed effects models where data from multiple buildings can be used to fit these relationships resulting in models that are more robust to outliers or sparse data at extremes.

@@

Unlike lagged demand variables, which are used to capture operational changes in a building, lagged temperature variables are used to capture thermal inertia. Thermal inertia is residual heat energy that remains in a building after a run of warm weather (or conversely for cold water). For example, if several warm days occur sequentially, the expected demand can increase as more cooling is typically required to maintain indoor environment quality. Hence, lagged temperature variables are based off all days - not just working days. Monday's lagged temperature variables will include temperatures observed over the weekend.

@@

(ref:temp-wh-scatter) Relationship between temperature and normalised electricity demand of two Australian office buildings during Summer. Logged values are shown as we use these as our response variable when fitting models to enforce a positivity constraint. Data at midday and midnight. The relationship between temperature and demand is different both between buildings and at different times of the day.

@@

ggplot(mapping = aes(x = temperature)) +
  geom_point(data = p_data, aes(y = log(wh)), shape = "O", alpha = 0.5) +
  geom_line(
    data = p_data %>%
      select(
        temperature, `Linear spline`, `Cubic spline`,
        `Natural spline`
      ) %>%
      gather(var, val, -temperature) %>%
      mutate(var = factor(var, levels = c("Linear spline", "Cubic spline", "Natural spline"))),
    aes(y = val, colour = var)
  ) +
  facet_wrap(~var) +
  labs(
    colour = "Method",
    y = expression(Log ~ normalised ~ electricity ~ (log(Wh / m^2))),
    x = "Temperature (°C)"
  ) +
  theme(legend.position = "none")
```

@@

Table: (\#tab:model-frameworks) Models

Model                   | Abbreviation | Description              | Weather Variables
------------------------|------------|--------------------------|-------------------
Naive                        | Naive | Naive forecasting model | None
Individual linear regression | ILR  | Linear regression models fit to each building | Current temperature
Individual natural splines   | INS  | Natural spline models fit to each building | Current temperature, selected features
Pooled regression            | PR | Regression model. Used for feature selection | Current temperature, selected features
Random intercept             | RI   | Mixed effects model with random intercept | Current temperature, selected features
Random intercept and slope   | RIS   | Mixed effects model with random intercept and slope | Current temperature, selected features
Subject-specific curves      | SSC  | Mixed effects model with subject-specific curves | Current temperature, selected features
Subject-specific curves with attributes | SSCATTR | SSC with building attributes included as fixed effects | Current temperature, selected features

Due to the evolving nature of energy demand across the day we fit separate models for each 15-minute period of the day. This gives 96 models for each building (when fitting individual models) and 96 models for each mixed effects formulation.

@@

A linear relationship between temperature and electricity demand may not be sufficient to adequately capture the relationship between the two. Natural cubic splines allow a more flexible relationship between predictors and the response. In this model the log demand of each building is modelled separately using natural splines. A building's demand based on temperature and other selected predictors is given by
$$
\begin{gathered}
\log y_{t} = f_p(w_{0,t}) + \epsilon_{t}, \quad \epsilon_{t} \sim N(0, \sigma^2_{p}), \\
f_p(x) = \sum^K_{k=1} \beta_{i,p,k} (x - \kappa_k)_+^3,
\end{gathered}
$$
where $x_{i,t}$ is the value of the $i$^th^ predictor variable at time $t$ and $f_{i,p}$ is a smooth function modelling the relationship between $x_{i,t}$ and the logged demand for period $p$. We refer to this set of models as the "Individual Natural Spline" (INS) models.

@@

Additional predictor variables are introduced in this model, such as lagged temperature variables; maximum, minimum and average temperatures; and lagged demand. A description of each predictor is presented in Table \@ref(tab:weather-vars). We denote this set of predictor variables as $\mathcal{P}_t$, which contains the selected variables for the 15-minute period of day and month at time $t$. The exact combination of variables is chosen via our feature selection methodology described in Section \@ref(feature-selection). This model is used when selecting features as it is much faster to train via OLS than it is to train a mixed effects model.

The demand of building $b$ at time $t$ is given by
$$
\begin{gathered}
\log y_{b,t} = \sum_{x_{b,i,t} \in \mathcal{P}_t } f_{i,p}(x_{b,i,t}) + \beta_{b,p}+ \epsilon_{b,t}, \quad \epsilon_{b,t} \sim N(0, \sigma^2_{p}), \\
f_{i,p}(x) = \sum^K_{k=1} \beta_{i,p,k} (x - \kappa_k)_+^3,
\end{gathered}
$$
where variables are defined as for the INS model. An additional fixed effect $\beta_{b,p}$ has been added to account for differences in each building's consumption. This effectively acts as a dummy variable. We call this our pooled regression (PR) model.

@@

The simplest mixed effects model is a random intercept (RI) model. We model the log of the demand by
$$
\begin{gathered}
\log y_{b,t} = \sum_{x_{b,i,t} \in \mathcal{P}_t } f_{i,p}(x_{b,i,t}) + u_{b,p} + \epsilon_{b,t}, \\
f_{i,p}(x) = \sum^K_{k=1} \beta_{i,p,k} (x - \kappa_k)_+^3, \\
\epsilon_{b,t} \sim N(0, \sigma^2_{\epsilon, p}), \quad u_{b,p} \sim N(0, \sigma_{u,p}^2),
\end{gathered}
$$
where $u_{b,p}$ is a random effect that controls the intercept of the model. This is similar in form to the pooled regression model, with the coefficient $\beta_{b,p}$ replaced by the random intercept $u_{b,p}$. We don't use this model for feature selection in Section \@ref(feature-selection) as it takes much longer to fit than the pooled regression model.

@@

We fit models to `r n_buildings` buildings across Australia. Separate models for each building were fitted as a benchmark. The overall predictive power of several mixed effects models were assessed against this benchmark. One-day ahead forecasts were produced for business days over a year using all forecast methods. Based on the MAE, MAPE, sMAPE and MASE scores of each model the SSC and SSCATTR models performed best. We concluded that moving towards a mixed model approach improved predictive power and also had the additional advantage of allowing us to include fixed effects which could be used for scenario analyses.

Finally, we included an example of using the SSCAR model for scenario analysis for one of the buildings in our data set. The expected change in electricity demand was plotted when several of the attributes were varied. The expected change in consumption over the course of one workweek when a building moved away from using a DX system allowed us to quantify the potential savings in energy. Analyses such as these have applications for decision makers and facility managers that wish to understand the effectiveness of changes in building management or potential retrofits of equipment.

>>>>

---
title: "Subject-specific curves for time series forecasting of smart meter demand"
author:
- familyname: Roach
  othernames: Cameron
  address: Monash University
  email: cameron.roach@monash.edu
  correspondingauthor: true
- familyname: Hyndman
  othernames: Rob
  address: Monash University
- familyname: Taieb
  othernames: Souhaib Ben
  address: University of Mons
abstract: Buildings are typically equipped with smart meters to measure electricity demand at regular intervals. Smart meter data for a single building have many uses, such as forecasting and assessing overall building performance. However, when data are available from multiple buildings, there are additional applications that are rarely explored. For instance, we can explore how different building characteristics influence energy demand. If each building is treated as a random effect and building characteristics are handled as fixed effects, a mixed effects model can be used to estimate how characteristics affect energy usage. In this paper we demonstrate that producing one-day ahead demand predictions for 123 commercial office buildings using mixed models can improve forecasting accuracy. We experiment with random intercept, random intercept and slope, and subject-specific curve mixed models. The predictive performance of the mixed effects models are tested against naive, linear and nonlinear benchmark models fitted to each building separately. Having justified the use of a mixed model framework, we provide an example showing how mixed model frameworks can, when combined with smart meter data and building attributes, be used to carry out scenario analysis. We demonstrate how expected electricity consumption may increase or decrease given a change in building attributes. This research justifies using mixed models to improve forecasting accuracy and to quantify changes in energy consumption under different building configuration scenarios.
keywords: "time series forecasting, mixed-effects models, smart meters, energy, electricity"
wpnumber: 1/19
jelcodes: C10,C14,C52
blind: false
cover: false
toc: false
bibliography: ["library.bib", "packages.bib"]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    # includes:
    #   in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

To the author's knowledge few papers have explored using mixed models in an electricity demand forecasting role. @Brabec2008-jf appears to be closest to this area. In their paper, a nonlinear mixed effects model (NLME) was used to forecast daily gas demand for individual customers. Predictors such as day of week and temperature were treated as random effects. Their NLME model was benchmarked against ARIMAX and ARX approaches. The paper concluded by saying there was no clear winner between the NLME and benchmark models and that both potentially have strengths and weaknesses. Unfortunately, there are few other papers within the energy field that use mixed effects models^[Some papers claim to use mixed models. However, this term is often applied to cases where a combination of models have been used which is different to mixed effects models in the statistical sense.] for forecasting.

@@

Few papers have attempted to assess the impact of differences in building characteristics using statistical methods and smart meter data. To the author's knowledge, only a previous paper by @Roach2019-ss has looked into this using mixed effects models. Whereas that paper focused on estimating demand impact profiles for building attributes at different times of the year, this paper focuses on improving forecast accuracy and producing scenario analyses by estimating the expected change in time series conditional on the building attributes.

@@

The main contribution of this paper is to present an approach to forecasting electricity demand for individual buildings using a mixed effects framework. Furthermore, we show how such a model can be used to conduct scenario analysis allowing us to quantify expected energy savings given changes in building attributes. Finally, this paper serves to enrich the literature on forecasting with mixed effects models and smart meter data.

The paper is structured as follows. Section \@ref(data) describes the data we are working with. Section \@ref(methodology) gives a detailed description of the models formulations and how they are assessed. Forecasting results and a scenario analysis example are presented in Section \@ref(results). Concluding remarks are given in Section \@ref(conclusion).

@@

The relationship between current temperature and electricity demand is shown in Figure \@ref(fig:temp-wh-scatter) for two buildings at midday and midnight. There is a clear difference in this relationship between the business and non-business periods. Furthermore, this relationship varies between each building. Note that when splines are fitted independently for each building they can be very sensitive to outliers or data at domain extremes. This partly motivates our use of mixed effects models where data from multiple buildings can be used to fit these relationships resulting in models that are more robust to outliers or sparse data at extremes.

@@

Unlike lagged demand variables, which are used to capture operational changes in a building, lagged temperature variables are used to capture thermal inertia. Thermal inertia is residual heat energy that remains in a building after a run of warm weather (or conversely for cold weather). For example, if several warm days occur sequentially, the expected demand can increase as more cooling is typically required to maintain indoor environment quality. Hence, lagged temperature variables are based off all days - not just working days. Monday's lagged temperature variables will include temperatures observed over the weekend.

@@

(ref:temp-wh-scatter) Relationship between temperature and normalised electricity demand at midday and midnight for two Australian office buildings during Summer. Logged values are shown as we use these as our response variable when fitting models to enforce a positivity constraint. The relationship between temperature and demand is different both between buildings and at different times of the day.

@@

ggplot(mapping = aes(x = temperature)) +
  geom_point(data = p_data, aes(y = log(wh)), shape = "O", alpha = 0.5) +
  geom_line(
    data = p_data %>%
      select(
        temperature, `Linear spline`, `Cubic spline`,
        `Natural spline`
      ) %>%
      gather(var, val, -temperature) %>%
      mutate(var = factor(var, levels = c("Linear spline", "Cubic spline", "Natural spline"))),
    aes(y = val), colour = ba_palette[1]
  ) +
  facet_wrap(~var) +
  labs(
    colour = "Method",
    y = expression(Log ~ normalised ~ electricity ~ (log(Wh / m^2))),
    x = "Temperature (°C)"
  ) +
  theme(legend.position = "none")
```

@@

Table: (\#tab:model-frameworks) Model descriptions

Model                   | Abbreviation | Description              | Predictor Variables
------------------------|------------|--------------------------|-------------------
Naive                        | Naive | Naive forecasting model  | Previous day's demand
Individual linear regression | ILR   | Linear regression models fit to each building | Current temperature
Individual natural splines   | INS  | Natural spline models fit to each building | Current temperature
Pooled regression            | PR | Regression model. Used for feature selection | Current temperature, selected features
Random intercept             | RI   | Mixed effects model with random intercept | Current temperature, selected features
Random intercept and slope   | RIS   | Mixed effects model with random intercept and slope | Current temperature, selected features
Subject-specific curves      | SSC  | Mixed effects model with subject-specific curves | Current temperature, selected features
Subject-specific curves with attributes | SSCATTR | SSC with building attributes included as fixed effects | Current temperature, selected features

Due to the evolving nature of energy demand across the day we fit separate models for each 15-minute period of the day. This gives 96 models for each building when fitting individual model formulations and 96 models for each mixed effects formulation.

@@

A linear relationship between temperature and electricity demand may not be sufficient to adequately capture the relationship between the two. Natural cubic splines allow a more flexible relationship between predictors and the response. In this model the log demand of each building is modelled separately using natural splines. A building's demand based on temperature and other selected predictors is given by
$$
\begin{gathered}
\log y_{t} = f_p(w_{0,t}) + \epsilon_{t}, \quad \epsilon_{t} \sim N(0, \sigma^2_{p}), \\
f_p(x) = \sum^K_{k=1} \beta_{p,k} (x - \kappa_k)_+^3,
\end{gathered}
$$
where $f_p$ is a smooth function modelling the relationship between $w_{0,t}$ and the logged demand for period $p$. We refer to this set of models as the "Individual Natural Spline" (INS) models.

@@

Additional predictor variables are introduced in this model, such as lagged temperature variables; maximum, minimum and average temperatures; and lagged demand. A description of each predictor is presented in Table \@ref(tab:weather-vars). We denote this set of predictor variables as $\mathcal{P}_t$, which contains the selected variables for the 15-minute period of day and month at time $t$. The exact combination of variables is chosen via our feature selection methodology described in Section \@ref(feature-selection). This model is used when selecting features as it is much faster to train than a mixed effects model.

The demand of building $b$ at time $t$ is given by
$$
\begin{gathered}
\log y_{b,t} = \sum_{x_{b,i,t} \in \mathcal{P}_t } f_{i,p}(x_{b,i,t}) + \alpha_{b,p}+ \epsilon_{b,t}, \quad \epsilon_{b,t} \sim N(0, \sigma^2_{p}), \\
f_{i,p}(x) = \sum^K_{k=1} \beta_{i,p,k} (x - \kappa_k)_+^3,
\end{gathered}
$$
where $x_{b,i,t}$ is the value of building $b$'s $i$^th^ predictor variable at time $t$ and $f_{i,p}$ is a smooth function modelling the relationship between $x_{b,i,t}$ and the logged demand for period $p$. A dummy variable $\alpha_{b,p}$ has been added to account for differences in each building's consumption. We call this our "Pooled Regression" (PR) model.

@@

The simplest mixed effects model is a random intercept (RI) model. We model the log of the demand by
$$
\begin{gathered}
\log y_{b,t} = \sum_{x_{b,i,t} \in \mathcal{P}_t } f_{i,p}(x_{b,i,t}) + u_{b,p} + \epsilon_{b,t}, \\
f_{i,p}(x) = \sum^K_{k=1} \beta_{i,p,k} (x - \kappa_k)_+^3, \\
\epsilon_{b,t} \sim N(0, \sigma^2_{\epsilon, p}), \quad u_{b,p} \sim N(0, \sigma_{u,p}^2),
\end{gathered}
$$
where $u_{b,p}$ is a random effect that controls the intercept of the model. This is similar in form to the pooled regression model, with the dummy variable $\alpha_{b,p}$ replaced by the random intercept $u_{b,p}$. We don't use this model for feature selection in Section \@ref(feature-selection) as it takes much longer to fit than the pooled regression model.

@@

We fit models to `r n_buildings` buildings across Australia. Separate models for each building were fitted as a benchmark. The overall predictive power of several mixed effects models were assessed against this benchmark. One-day ahead forecasts were produced for business days over a year using all forecast methods. Based on the MAE, MAPE, sMAPE and MASE scores of each model the SSC and SSCATTR models performed best. We concluded that predicting demand using mixed effects model could improve forecast accuracy and had the additional advantage of allowing us to conduct scenario analyses.

Finally, we included an example of using the SSCATTR model for scenario analysis for one of the buildings in our data set. The expected change in electricity demand was plotted when several of the attributes were varied. The expected change in consumption over the course of one workweek when a building moved away from using a DX system allowed us to quantify the potential savings in energy. Analyses such as these have applications for decision makers and facility managers that wish to understand the effectiveness of changes in building management or potential retrofits of equipment.
